---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

# medmodr Package: Iterated Moderation and Mediation Analysis at Scale

<!-- badges: start -->
[![R-CMD-check](https://github.com/tosmartak/medmodr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/tosmartak/medmodr/actions/workflows/R-CMD-check.yaml)
[![License: GPL v3](https://img.shields.io/badge/License-GPL%20v3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)
<!-- badges: end -->

Systematic moderation and mediation analysis in R. 

`medmodr` helps you scan multiple variable combinations for moderation and mediation effects without having to manually fit each model. It provides tidy outputs and ready-to-publish visualizations. This package builds upon the [`mediation`](https://CRAN.R-project.org/package=mediation) package [Tingley et al., 2014] for mediation analysis and the [`interactions`](https://CRAN.R-project.org/package=interactions) package [Long, 2024] for moderation plots.

## Learn more
Full walkthrough with moderation, mediation, and plotting:
[`Getting started with medmodr`](https://tosmartak.github.io/medmodr/articles/getting-started.html)

## Installation

`medmodr` requires `R >= 4.1.0`.

You can install the released version of **medmodr** from CRAN with:

```{r install-cran, eval=FALSE}
install.packages("medmodr")
```

And you can install the development version from GitHub with:

```{r install-github, eval=FALSE}
# install.packages("pak")
pak::pak("tosmartak/medmodr")
# or alternatively:
# devtools::install_github("tosmartak/medmodr")
```

```{r setup-example, message=FALSE}
library(medmodr)
```

## Example
The package includes a small demo dataset `demo_medmodr`. Below we show one minimal moderation and mediation analysis.

```{r load-dataset}
data("demo_medmodr")
```

### Moderation example

```{r run-moderation}
mod_summary <- run_moderation_paths(
  data = demo_medmodr,
  predictors = c("x1", "x2"),
  moderators = c("m1", "grp", "edu"), # mix of numeric and categorical
  outcomes = c("y1", "y2"),
  controls = c("c1", "c2"),
  categorical_vars = c("grp", "edu"),
  sig_level = 0.05,
  plot_sig = FALSE, # set TRUE in your analysis to auto-plot significant interactions using the interaction package
  summarize_categorical = TRUE # summarize multiple dummy-by-interaction lines into a single row per categorical mod
)
```

The resulting dataset from our moderation analysis is shown below:

```{r view-moderation-dataset}
knitr::kable(head(mod_summary))
```

### Mediation example

```{r run-mediation-analysis}
med_summary <- run_mediation_paths(
  data = demo_medmodr,
  treatments = c("x1", "x2"), # you can also include "grp" if you want to treat it as a treatment
  mediators = c("m1", "m2"),
  outcomes = c("y1", "y2"),
  controls = c("c1", "c2", "edu"),
  sims = 200, boot = FALSE, seed = 1
)
```

The resulting dataset from our mediation analysis is shown below:

```{r view-mediation-dataset}
knitr::kable(head(med_summary))
```

#### Visual summaries for mediation
These plots are fast for small tables but can be heavy on large scans, so we kept most of them eval=FALSE in README. Use them interactively in your analysis.

**Effect overview plot (ACME, ADE, Total Effect with CIs):**

By default, it would generate a summary grid plot for significant mediations only

```{r plot-med-summary-plot, eval=FALSE}
plot_mediation_summary_effects(med_summary, filter_significant = TRUE, summary_plot = TRUE)
```

**You can also show only ACME plots from significant results**
```{r, eval=FALSE}
plot_mediation_summary_effects(med_summary, filter_significant = TRUE, show_only_acme = TRUE, summary_plot = TRUE)
```

**You can equally loop over all significant mediation triples and plot each**

```{r, eval=FALSE}
plot_mediation_summary_effects(med_summary, filter_significant = TRUE, summary_plot = FALSE)
```

**You can also show only ACME for single plots**

```{r, eval=FALSE}
plot_mediation_summary_effects(med_summary, filter_significant = TRUE, summary_plot = FALSE, show_only_acme = TRUE)
```

## Function reference (quick)

- `run_moderation_paths(data, predictors, moderators, outcomes, controls, categorical_vars = NULL, sig_level = 0.05, plot_sig = FALSE, summarize_categorical = FALSE)` — loop over predictors × moderators × outcomes

- `run_mediation_paths(data, treatments, mediators, outcomes, controls, sims = 1000, boot = TRUE, seed = 123)` — loop over treatments × mediators × outcomes

- `plot_mediation_summary_effects(summary_table, filter_significant = FALSE, show_only_acme = FALSE)` — visualize significant mediation results

## License and issues

- [![License: GPL v3](https://img.shields.io/badge/License-GPL%20v3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)

- Please file issues and feature requests on GitHub

- Contributions welcome (add tests where possible)

## Citation

If you use medmodr in your research, please cite it:

```{r, citation}
citation("medmodr")
```

